Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-19 doc/mc-paper.tex
Updates

\usepackage{amsmath}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{hyperref}

\newcommand{\pand}{\mathbin{\land^{\sf p}}}
\newcommand{\por}{\mathbin{\lor^{\sf p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^{\sf p}}
\DeclareMathOperator*{\Por}{\bigvee^{\sf p}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}

@@

%\nolinenumbers %uncomment to disable line numbering

@@

We have developed a program that generates POG representations from
graphs 
produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
find an assignment satisfying $\phi$ or generate a proof that no
such assignment exists.  They have applications across a variety of
domains including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond Boolean satisfiability.  For example, the {\em model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

@@

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the {\em deterministic
decomposable, negation normal form}
(det-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}
represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that proof production capabilites allow SAT solver developers to quickly
detect and diagnose bugs in their programs. This, in turn, has led
to more reliable SAT solvers.

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a general form that can encode all of the representations produced by current knowledge
compilers. The CPOG (for ``certified'' POG) file format then
captures both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF).  A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that, when satisified by a CPOG file, guarantee that it
encodes a well-formed POG and provides a valid equivalence proof.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a {\em decision} DNNF (dec-DNNF) representation (a restricted form of dec-DNNF),
and the {\em proof
  generator} uses this to generate a CPOG file.
The {\em proof checker} verifies the equivalence of the CNF and CPOG representations.
The {\em ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CPOG proof system, as well as correctness of its
concrete implementation, we formally verified the proof system as well as versions
of the proof checker and ring evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a particular CPOG file gives assurance that the proof
and the model count are correct. Our experience with developing a formally verified
proof checker has shown that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of requirements that guarantee soundness.
In fact, subtle details in the partitioned sum rule were identified during this process.

We evaluate our toolchain using benchmark
formulas from the 2022 standard and weighted model competitions.  Our
tools handle all but the largest dec-DNNF graphs generated
by \dfour{}.  We also measure the benefits of several optimizations as
well as the relative performance of the verified checker with one
designed for high performance and capacity.

@@

Capelli et al.\ developed a version of the \dfour{} knowledge compiler
that generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their
modified version of \dfour{} generates additional annotations to the
compiled representation, as well as a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
These files must be processed by two different checkers.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.  They
developed a program that can generate proof traces from dec-DNNF
graphs and a program to check adherence to their proof rules.  This
framework is not directly comparable to ours, since it only certifies
the unweighted model count, but it has similar goals.

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.  Although the authors make informal arguments
regarding the soundness of their frameworks, these do not provide
strong levels of assurance.  Indeed, as we discuss in
Appendix~\ref{app:comparison}, we uncovered a weakness in the \dfour{}
verification framework due to its invalid assumption about the
guarantees provided by \dtrim{}, the program it uses to check their
DRAT file.  Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In Appendix~\ref{app:comparison}
we also compare the performance of our toolchain to these two others.

@@

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an {\em assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
  \subseteq \varset$.  We say the assignment is {\em total} when it
  assigns a value to every variable ($\varset' = \varset$), and that
  it is {\em partial} otherwise.
  The set of all possible total assignments over
  $\varset$ is denoted $\uassign$.

For each variable $x \in \varset$,
  we define the {\em literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated {\em dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its {\em negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their {\em sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in {\em negation normal form} when negation is applied only to variables.  A
  formula is in {\em conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of {\em clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a {\em unit} clause and the literal as a {\em unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\begin{definition}[Commutative Ring]
  A {\em commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  {\em additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}
We write $a - b$ as a shorthand for $a + -b$.

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a {\em ring weight function} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\phi$ and ring weight function $w$, the {\em ring evaluation problem} computes
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) {\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

The {\em weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The {\em function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, chose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

@@

(D) Implicit Clauses\\[1.2ex]

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{tikz}
\usepackage{pgfplots}
\usepackage{booktabs}
\usepackage{hyperref}

\newcommand{\pand}{\mathbin{\land^\textsf{p}}}
\newcommand{\por}{\mathbin{\lor^\textsf{p}}}
\DeclareMathOperator*{\Pand}{\bigwedge^\textsf{p}}
\DeclareMathOperator*{\Por}{\bigvee^\textsf{p}}
\newcommand{\boolnot}{\neg}
\newcommand{\tautology}{\top}
\newcommand{\nil}{\bot}
\newcommand{\obar}[1]{\overline{#1}}
\newcommand{\oneminus}{{\sim}}
\newcommand{\lit}{\ell}

\newcommand{\varset}{X}
\newcommand{\exvarset}{Z}
\newcommand{\dependencyset}{{\cal D}}
\newcommand{\litset}{{\cal L}}
\newcommand{\ring}{{\cal R}}
\newcommand{\dset}{{\cal A}}
\newcommand{\rep}{\textbf{R}}
\newcommand{\radd}{+}
\newcommand{\rmul}{\times}
\newcommand{\addident}{\textbf{0}}
\newcommand{\mulident}{\textbf{1}}
\newcommand{\imply}{\Rightarrow}
\newcommand{\ifandonlyif}{\Leftrightarrow}
%\newcommand{\drational}{\mathbb{Q}_{2,5}}
\newcommand{\drational}{\textbf{Q}_{2,5}}
\newcommand{\entail}{\vDash}
\newcommand{\entaildrat}{\entail_{\textrm{DRAT}}}

@@

\newcommand{\progname}[1]{\textsc{#1}}
\newcommand{\dfour}{\progname{D4}}
\newcommand{\cdfour}{\progname{CD4}}
\newcommand{\Dfour}{\progname{D4}}
\newcommand{\cadical}{\progname{CaDiCal}}
\newcommand{\dtrim}{\progname{drat-trim}}
\newcommand{\Dtrim}{\progname{Drat-trim}}

@@

\nolinenumbers %uncomment to disable line numbering

\EventEditors{Meena Mahajan and Friedrich Slivovsky}
\EventNoEds{2}
\EventLongTitle{26th International Conference on Theory and Applications of Satisfiability Testing (SAT 2023)}
\EventShortTitle{SAT 2023}
\EventAcronym{SAT}
\EventYear{2023}
\EventDate{July 4--8, 2023}
\EventLocation{Alghero, Italy}
\EventLogo{}
\SeriesVolume{271}
\ArticleNo{6}

@@

We have developed a program that generates POG representations from det-DNNF
graphs 
produced by the state-of-the-art knowledge compiler
\dfour{}, as well as checkable CPOG proofs certifying that the output POGs
are equivalent to the input CNF formulas.  Our toolchain
for generating and verifying POGs scales to all but the largest
graphs produced by \dfour{} for formulas from a recent model counting
competition. Additionally, we have developed a formally verified CPOG
checker and model counter for POGs in the \lean{} proof assistant.
In doing so, we proved the soundness of our proof framework. These programs
comprise the first formally verified toolchain for weighted and unweighted
model counting.
\end{abstract}

@@

Given a Boolean formula $\phi$, modern Boolean satisfiability (SAT) solvers can
find an assignment satisfying $\phi$ or generate a proof that no
such assignment exists.  They have applications across a variety of
domains including computational mathematics, combinatorial
optimization, and the formal verification of hardware, software, and
security protocols.  Some applications, however, require going
beyond Boolean satisfiability.  For example, the \emph{model
  counting problem} requires computing the number of satisfying
assignments of a formula, including in cases where there are far too many
to enumerate individually.  Model counting has
applications in artificial intelligence, computer security, and
statistical sampling.  There are also many useful extensions of standard model counting,
including {\em
  weighted model counting}, where a weight is defined for
each possible assignment, and the goal becomes to compute the sum of the weights
of the satisfying assignments.

@@

One approach to model counting, known as \emph{knowledge compilation},
transforms the formula into a structured form for which model counting
is straightforward.  For example, the \emph{deterministic
decomposable, negation normal form}
(det-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004}
represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by a Boolean And or Or operation.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig et al.\ present a very general {\em
  algebraic model counting}~\cite{kimmig:jal:2017} framework describing
properties of Boolean functions that can be efficiently computed from
a det-DNNF representation.  These include standard and weighted model
counting, and much more.

One shortcoming of existing knowledge compilers is that they have no
generally accepted
way to validate that their generated results are correct, i.e., that
the compiled representation is logically equivalent to the original
formula.  By contrast, all modern SAT solvers can generate
checkable proofs when they encounter unsatisfiable formulas.  The
guarantee provided by a checkable certificate of correctness enables
users of SAT solvers to fully trust their results.  Experience has also
shown that proof production capabilites allow SAT solver developers to quickly
detect and diagnose bugs in their programs. This, in turn, has led
to more reliable SAT solvers.

This paper introduces \emph{Partitioned-Operation Graphs} (POGs),
a general form that can encode all of the representations produced by current knowledge
compilers. The CPOG (for ``certified'' POG) file format then
captures both the structure of a POG
and a checkable proof of its logical equivalence to a Boolean formula in
conjunctive normal form (CNF).  A CPOG
proof consists of a sequence of clause addition and deletion steps,
based on an extended resolution proof system~\cite{Tseitin:1983}.
We establish a set of conditions that, when satisified by a CPOG file, guarantees that it
encodes a well-formed POG and provides a valid equivalence proof.

@@

Figure~\ref{fig:chain} illustrates our certifying knowledge
compilation and model counting toolchain.  Starting with
input formula $\inputformula$, the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017}
generates a \emph{decision} DNNF (dec-DNNF) representation (a restricted form of det-DNNF),
and the \emph{proof
  generator} uses this to generate a CPOG file.
The \emph{proof checker} verifies the equivalence of the CNF and CPOG representations.
The \emph{ring evaluator} computes
a standard or weighted model count from the POG\@ representation.
As the dashed box in Figure~\ref{fig:chain} indicates, this toolchain
moves the root of trust away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will be caught by the proof checker.

To ensure soundness of the abstract CPOG proof system, as well as
correctness of its concrete implementation, we formally verified the
proof system as well as versions of the proof checker and ring
evaluator in the \lean{} proof assistant~\cite{demoura:cade:2021}.
Running these two programs on a particular CPOG file gives strong
assurance that the proof and the model count are correct. Our
experience with developing a formally verified proof checker has shown
that, even within the well-understood framework of extended
resolution, it can be challenging to formulate a full set of
requirements that guarantee soundness.  In fact, our efforts to
formally verify our proof framework exposed subtle conditions that we had
to impose on our partitioned sum rule.

We evaluate our toolchain using benchmark formulas from the 2022
standard and weighted model competitions.  Our tools handle all but
the largest dec-DNNF graphs generated by \dfour{}.  We also measure
the benefits of several optimizations as well as the relative
performance of the verified checker with one designed for high
performance and capacity.  We also show that our tools can provide
end-to-end verification of formulas that have been transformed by an
equivalence-preserving preprocessor.

@@

Capelli et al.\ developed a knowledge compiler
that generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{} program,
a modified version of \dfour{}, generates annotations to the
compiled representation, providing information about how the compiled version relates to the input clauses.
It also generates a file of clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
Completing the certification involves running two different checkers on
the annotated dec-DNNF graph and the DRAT file.
Although the authors make informal arguments
regarding the soundness of their frameworks, these do not provide
strong levels of assurance.  Indeed, 
we have identified a weakness in the their
methodology due to an invalid assumption about the
guarantees provided by \dtrim{}, the program it uses to check the
DRAT file.

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi$, their first task is to
check whether $\phi \imply R$, i.e., that any model of $\phi$ is also
a model of $R$.  They then base other parts of their proof on that
property and use a separate checker to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it may
not preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_1 \lor \obar{x}_2)$\\    
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has five.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however the
second clause of $\phi_2$ is RAT with respect to the first---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2 = 0$.

This weakness would allow a buggy (or malicious) version
of \cdfour{} to generate a compiled representation of $\phi_2$ when
given $\phi_1$ as its input.  By adding the second clause to $R$,
the check with \dtrim{} would pass, as would the other tests performed
by their checker.  We have been able to confirm this possibility with
their compiler and checker\footnote{Downloaded May 18, 2023 from
\url{https://github.com/crillab/d4}}.

This weakness can be corrected by restricting \dtrim{} to only add
clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
commandline argument to \dtrim{}\footnote{Available at
\url{https://github.com/marijnheule/drat-trim/tree/develop}.} that enforces this
restriction.  This weakness, however, illustrates the general challenge of
developing a new proof framework.
Without a formal verification, there are likely to be 
conditions that make the framework unsound.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithms commonly used by model counters.  They
developed a program that can generate proof traces from dec-DNNF
graphs and a program to check adherence to their proof rules.  This
framework is not directly comparable to ours, since it only certifies
the unweighted model count, but it has similar goals.
Again, they provide only  informal arguments 
regarding the soundness of their frameworks.

Both of these prior certification frameworks are strongly tied to the
algorithms used by the knowledge compilers and model counters.  Some
of the conditions to be checked are relevant only to specific
implementations.    Our framework is very general and is based on a small set
of proof rules.  It builds on the highly developed
concepts of clausal proof systems.  These factors were important in enabling formal verification.
In Appendix~\ref{app:comparison}
we also compare the performance of our toolchain to these two others.

@@

  Let $\varset$ denote a set of Boolean variables, and let $\assign$
  be an \emph{assignment} of truth values to some subset of the
  variables, where $0$ denotes false and $1$ denotes true, i.e.,
  $\assign \colon \varset' \rightarrow \{0,1\}$ for some $\varset'
  \subseteq \varset$.  We say the assignment is \emph{total} when it
  assigns a value to every variable ($\varset' = \varset$), and that
  it is \emph{partial} otherwise.
  The set of all possible total assignments over
  $\varset$ is denoted $\uassign$.

For each variable $x \in \varset$,
  we define the \emph{literals} $x$ and $\obar{x}$, where $\obar{x}$ is the
  negation of $x$. An
  assignment $\assign$ can be viewed as a set of literals, where
  we write $\lit \in \assign$ when $\lit = x$ and $\assign(x) = 1$ or when
  $\lit = \obar{x}$ and $\assign(x) = 0$.  We write the negation of literal $\lit$ as $\obar{\lit}$.  That is, $\obar{\lit} = \obar{x}$ when $\lit = x$ and
$\obar{\lit} = x$ when $\lit = \obar{x}$.

\begin{definition}[Boolean Formulas]
  The set of Boolean formulas is defined recursively.  Each
  formula $\phi$ has an associated \emph{dependency set}
  $\dependencyset(\phi)  \subseteq \varset$, and a set of models $\modelset(\phi)$,
  consisting of total assignments that satisfy the formula:
  \begin{enumerate}
  \item Boolean constants $0$ and $1$ are Boolean formulas,
    with $\dependencyset(0) = \dependencyset(1) = \emptyset$, with $\modelset(0) = \emptyset$, and with $\modelset(1) = \uassign$.
  \item Variable $x$ is a Boolean formula, with $\dependencyset(x) = \{x\}$
    and $\modelset(x) = \{\assign \in \uassign | \assign(x)=1\}$.
  \item For formula $\phi$, its \emph{negation}, written $\boolnot \phi$ is a Boolean formula,
    with $\dependencyset(\boolnot \phi) = \dependencyset(\phi)$ and $\modelset(\boolnot \phi) = \uassign - \modelset(\phi)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{product} $\phi = \bigwedge_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcap_{1 \leq i \leq k} \modelset(\phi_i)$.
  \item For formulas $\phi_1, \phi_2, \ldots, \phi_k$, their \emph{sum} $\phi = \bigvee_{1 \leq i \leq k} \phi_i$ is a Boolean formula, with
      $\dependencyset(\phi) = \bigcup_{1 \leq i \leq k} \dependencyset(\phi_i)$ and
      $\modelset(\phi) = \bigcup_{1 \leq i \leq k} \modelset(\phi_i)$.
  \end{enumerate}
\label{def:boolean}
\end{definition}

  We highlight two special classes of Boolean formulas.  A formula is
  in \emph{negation normal form} when negation is applied only to variables.  A
  formula is in \emph{conjunctive normal form} (CNF) when i) it is in
  negation normal form, and ii) sum is applied only to literals.  A CNF
  formula can be represented as a set of \emph{clauses}, each of which is a
  set of literals.  Each clause represents the sum of the
  literals, and the formula is the product of its clauses.  We use
  set notation to reference the clauses within a formula and the
  literals within a clause.  A clause consisting of a single literal is referred to as a \emph{unit} clause and the literal as a \emph{unit} literal.
This literal must be assigned value $1$ by any satisfying assignment of the formula.

\begin{definition}[Partitioned-Operation Formula]\label{def:partitioned-operation-formula}
  A \emph{partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\begin{definition}[Commutative Ring]
  A \emph{commutative ring} $\ring$ is an algebraic structure
  $\langle \dset, \radd, \rmul, \addident, \mulident \rangle$,
  with elements in the set $\dset$ and with commutative and
  associative operations $\radd$ (addition) and $\rmul$ (multiplication),
  such that multiplication distributes
  over addition.  $\addident$ is the additive identity and $\mulident$ is
  the multiplicative identity.  Every element $a \in \dset$ has an
  \emph{additive inverse} $-a$ such that $a + -a = \addident$.
\label{def:ring}
\end{definition}
We write $a - b$ as a shorthand for $a + -b$.

\begin{definition}[Ring Evaluation Problem]
\label{def:ring_evaluation}
  For commutative ring $\ring$, a \emph{ring weight function} associates a value $w(x) \in \dset$ with
  every variable $x \in \varset$.  We then define $w(\obar{x}) \doteq \mulident-w(x)$.

  For Boolean formula $\phi$ and ring weight function $w$, the \emph{ring evaluation problem} computes
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:weight}
\end{definition}

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
(standard) \emph{model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
weight function $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the \emph{density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

The \emph{weighted model counting}  problem is also defined over
rational numbers.  Some formulations  allow
independently assigning weights $W(x)$ and $W(\obar{x})$ for each variable $x$ and its complement, with the possibility that
$W(x) + W(\obar{x}) \not = 1$.
We can cast this as a
ring evaluation problem by letting $r(x) = W(x) + W(\obar{x})$,
performing ring evaluation with weight function $w(x) = W(x)/r(x)$ for each
variable $x$, and computing the weighted count
as $\rep(\phi, w)\; \rmul\; \prod_{x \in \varset} r(x)$.
Of course, this requires that $r(x) \not = 0$ for all $x \in \varset$.

The \emph{function hashing problem} provides a test
of inequivalence for Boolean formulas.  That is, for $n = |\varset|$, let $\ring$ be a
finite  field with $|\dset| = m$ such that $m \geq 2 n$.  For each $x \in \varset$, choose a value from $\dset$ at random for $w(x)$.  Two formulas
$\phi_1$ and $\phi_2$ will clearly have $\rep(\phi_1, w) = \rep(\phi_2, w)$
if they are logically equivalent.
%, and if $\rep(\phi_1, w) \not = \rep(\phi_2, w)$, then they are clearly inequivalent.
If they are not equivalent, then
the probability that $\rep(\phi_1, w) \not = \rep(\phi_2, w)$ will be at
least $\left(1-\frac{1}{m}\right)^n \geq \left(1-\frac{1}{2n}\right)^n > 1/2$.
Function hashing can therefore be used as part of a
randomized algorithm for equivalence testing~\cite{blum:ipl:1980}.
For example, it can compare different runs on a single formula,
either from different compilers or from a single compiler with different configuration parameters.

@@

(D) Defining Clauses\\[1.2ex]

@@

\subsection{Certifying Preprocessed Formulas}
